Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-07-09 Delaneyetal2024.tex

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom.
  This quantity varies with water velocity and channel width.
  In subaerial channels,  sediment transport capacity can evolve consistently because water discharge variations are accommodated by changes in flow depth, width, and water velocity.
  By contrast, subglacial channels grow at a slower rate than most water discharge variations, causing a variable relationship between channel shape, water velocity, and discharge.
  Using numerical models applied to two observed hydrographs and formulations, we evaluate the differences in sediment transport capacity variability between subaerial and subglacial channels.
  We also establish the conditions under which subglacial sediment transport capacity could co-vary with water discharge.
  The formulations of hydraulics and numerical models confirm that subglacial sediment transport is highly non-linear with respect to water discharge.
  Yet, subglacial sediment transport capacity can vary less if water discharge variations are closer to equilibrium with subglacial channel size.
  In a dedicated numerical experiment, we show that the changing channel size results in sediment transport capacity peaking before the maximum water discharge.
  Results also indicate that high subglacial sediment transport capacities can occur across a wide range of water discharges.
  This implies that hysteresis in channel size causes a variable relationship between sediment and water discharge in a transport-limited subglacial system.
  The implications are discussed for sediment discharge from glaciers with different hydro-climatic forcings and assumptions of channel shape.
  These findings can also improve interpretations of sediment discharge records in glacierized catchments.
\end{abstract}

@@

This manuscript has two objectives: 1) to evaluate the variability of subglacial sediment transport capacity with evolving subglacial channel size and different hydrographs and 2) to establish whether sub-seasonal water discharge can co-vary with sediment transport capacity in subglacial systems.
First, to demonstrate subglacial sediment transport capacity's highly variable nature,  we present algebraic formulations of the sediment transport capacity response to water discharge in subaerial, pipe-flow, and equilibrium R-channel conditions.
These formulations extend the relationships presented in \citet{alley1997}.
We then use numerical models to examine sediment transport capacity in both subglacial channels that evolve in size and subaerial channels that remain fixed.
They are applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and a land-terminating glacier in Greenland (Leverett Glacier).
We run a model ensemble to evaluate the sediment transport capacity variability in subglacial channels compared to subaerial channels.
Model outputs demonstrate the specific processes that can drive variability in sediment discharge capacity in subglacial systems.
These findings indicate differences in the relationships amongst water discharge, channel geometry, and water velocity in subglacial and subaerial channels.
The manuscript then discusses the implications of variability in sediment transport capacity from glaciers and the interpretation of sediment transport records.

@@

\begin{table}[hbt!]
  \centering
  \caption{Variables, parameters, and constants used in this work.
    Where two values are given, the first refers to  \alpine{}, a scenario from Fieschergletscher, and the second to a glacier marginal to the \icesheet{} a scenario from Leverett Glacier.
    A second line refers to the range of values examined in the parameter search.}
  \small 
  \begin{tabular}{ l  c  c c }
    Name &Symbol&  Value&Units \\
         && (\alpine{} or \icesheet{})\\
    \hline
    \textbf{Variables}  & & & \\
    Water discharge  & $Q$& & $\mathrm{m^{3}\,s^{-1}}$ \\
    Water velocity (subglacial, subaerial)  & $v$, ($v_g,\,v_{f}$)& & $\mathrm{m\,s^{-1}}$ \\
    Channel wetted area (subglacial, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
    Channel depth (subaerial) & $H$&& $\mathrm{m}$\\
    Hydraulic diameter &$D_h$&&$\mathrm{m}$\\
    Width of channel floor (subglacial, subaerial) & $w$, ($w_g,w_f$)&  & $\mathrm{m}$     \\
    Hydraulic head &$\Delta h$&& $\mathrm{m}$\\
    Hydraulic gradient &$\Psi=\frac{\Delta h}{l}$&& $\mathrm{m\, m^{-1}}$\\
    Shear stress (subglacial, subaerial) & $\tau$, ($\tau_g,\,\tau_f$) && $\mathrm{Pa \, m^{-2}}$ \\
    Stream power & $\Omega$ && $\mathrm{ kg \, m\, s^{-3}}$ \\

@@

\section{Results}
%We then apply the models above (Section~\ref{sect:meth}) in two experiments by examining three model outputs.
%Water velocity and shear stress represent the potential for sediment mobilization, and width-integrated shear stress represents a proxy for the total sediment transport capacity across the channel bed (Figure~\ref{fig:model_outs}).

@@

While transport-limited states at glaciers are likely rare \citep[e.g.][]{alley1997}, abundant sediment could persist underneath some glaciers \citep[e.g.][]{walter2014,stevens2022,delaney2022}.
As a result, sediment export can be hydrologically controlled and efficient \citep[e.g.][]{swift2005,andresen2024}.
At these glaciers, the great variability in subglacial sediment transport capacity may make it difficult to link sediment discharge to hydrology, especially when peak events occur \citep{cowan1988,delaney2018,lu2022}.
As a result, the sediment transport capacity response to individual extreme precipitation or melt events could be difficult to establish.
The results here suggest that different channel sizes with the same hydrology forcing can result in very different sediment transport capacities.

Catchments with reduced water discharge variability may experience yet less variability in sediment transport capacity, shown by the \alpine{} and \icesheet{} model outputs.
Indeed, the more decoupled and sporadic relationship between model outputs and water discharge in \alpine{} results from the relatively larger discharge variations on sub-daily to weekly timescales (Figure~\ref{fig:Qw_vari}).
This occurs as the subglacial channel's size evolves slowly compared to the variations in water discharge.
High water discharge variability in \alpine{} may cause shear stress to approach $Q^{2}$ in assuming pipe-flow conditions where the water discharge varies largely compared to channel size \citep[Section \ref{sect:scaling}; c.f.][]{alley1997}.
Conversely, the reduced relative variability in water discharge in \icesheet{} and stronger correlation between width-integrated shear stress may result in the subglacial channel's size being closer to equilibrium from the smaller variations in water discharge.
In this case, the exponent on water discharge for shear stress is likely substantially less than $w\tau \propto Q^2$, but greater than $w \tau \propto Q^{\frac{4}{5}}$ that occurs in a steady-state R-channel (see Section \ref{sect:scaling}).
As a result, there is a stronger relationship between subglacial model outputs and water discharge in \icesheet{} (Figure~\ref{fig:Qw_vari}) and less variability model outputs in the \icesheet case (Figure~\ref{fig:multi_run}).

@@

The co-varying relationship between sediment discharge capacity and water discharge in subaerial conditions could be noticeable $\sim20$\,\unit{km} downstream of the Leverett site.
A strong correlation persists between sediment plume size and the Watson River's water discharge into the Kangerlussuaq fjord \citep{chu2009,mcgrath2010}.
In contrast, in marine-terminating glacier catchments, a less consistent relationship may occur between water discharge or melt extent and sediment plume size \citep{chu2012,tedstone2012}.
Here, ocean water pressurizes subglacial channels at the ice front \citep[e.g.][]{how2017}, so the observed reduced correlation could result from the inconsistent relationship between subglacial sediment transport capacity and water discharge (Figure~\ref{fig:Qw_vari}).

Water discharge measurements at less than daily timescales could severely limit subglacial sediment transport models in capturing specific events.
The decreased model output variability beyond $1$--$5$ days of smoothing could result in water discharge being in equilibrium with the subglacial channel when they are in fact not (Figure~\ref{fig:multi_run}).
Over this period, water discharge variations could lead to a stronger relationship with sediment discharge capacity (Section ~\ref{sect:scaling}).
Yet, such a strong relationship would not represent the impact of shorter-term fluctuations in water discharge.
Note, that if basal melt largely dominates, then water discharge would evolve slowly with respect to the subglacial channel size.

Clockwise hysteresis loops between sediment concentration and water discharge suggest that sediment availability is reduced over the event scale \citep[][]{williams1989}.
These loops are often observed in glacierized catchments and used to suggest that access to subglacial sediment has been limited \citep[e.g.][]{collins1979,willis1996,richards2003,stott2007,delaney2018}.
Results here suggest that clockwise hysteresis could also be expected in transport-limited regimes in many subglacial environments.
Here, the smaller channel size on the rising limb would result in greater sediment transport capacity and sediment uptakes.
A larger channel size would reduce sediment transport capacity for an equivalent water discharge on the falling limb, creating clockwise hysteresis.

@@

The manuscript's first objective aims to evaluate the relative variability in sediment transport capacity between the two channel types.
Results demonstrate that sediment transport capacity variability is higher in subglacial channels with evolving size across a range of channel conditions.
Increased variability occurs even in environments where the water discharge has relatively small diurnal variations in the \icesheet case.
Yet, elevated variability is especially strong with highly variable water discharge in the \alpine case.
Increased variability in sediment transport capacity in evaluating hydro-climatic signals from sediment records in glacierized catchments needs further evaluation.
Furthermore, greater variability in sediment transport capacity may lead to sediment exhaustion by subglacial water repeatedly crossing the mobilization threshold across a range of water discharges.

The manuscript's second objective is to establish if water discharge varies with subglacial sediment transport capacity.
Results here suggest that, even in a transport-limited subglacial system, an incoherent relationship between water and subglacial sediment discharge could be expected in most systems.
This presents a challenge in linking hydro-climatic conditions or events to sediment export from glaciers.
In subglacial channels, the timing of peak sediment discharge capacity occurs before peak water discharge during an event, due to evolving channel size.
In subaerial channels, the timing of peak sediment discharge capacity and water discharge coincide.
whereas it is coincident in subaerial ones.
Covariance between water discharge and sediment transport capacity could be possible when water discharge varies at a slower rate than subglacial channel size.
In natural environments, this appears to rarely be the case, except where basal melt dominates.

@@

\authorcontribution{
  I.D. designed the study, developed, and implemented the model and experiments, and wrote the manuscript.
  A.J.T. helped interpret findings from Leverett Glacier and guided experiment design.
  M.A.W. provided data from Fieschergletscher and added Section \ref{sect:scaling}.
  D.F. provided data from Fieschergletscher.
  All authors provided key inputs in writing and editing the manuscript.} %% this section is mandatory

@@

Here three kinds of drainage regimes are examined, namely, subaerial channels, steady-state R-channels \citep{rothlisberger1972}, and pipe-flow (i.e. R-channels which do not have time to adjust their size to discharge conditions).
Here we show the sediment transport capacity scaling with a given water discharge and hydraulic gradient for three different sediment transport capacity formulas: Meyer-Peter M\"uller  \citep[MPM; ][]{meyer1948}, Engelund and Hansen \citep[EH; ][]{engelund1967}, and Bagnold \citep{bagnold1980}; additionally width-integrated shear stress is assessed, as used as proxy for sediment transport throughout the manuscript.

Sediment discharge is given by the MPM, EH, and Bagnold formulations as
\begin{equation}
  Q_s \propto w\, \tau^{3/2}, \quad Q_s \propto w\, \tau^{5/2}, \quad Q_s \propto w \left(\frac{\Omega}{w}\right)^{3/2} H^{-2/3},
\end{equation}
respectively, for conditions well above the transport onset threshold.

@@

As in Section~\ref{sect:fluv}, the subaerial channel is assumed to have a width $w$ much greater than its depth $H$, such that $D_h\approx 4h$, and to have a constant head gradient ($\Psi$) given by the topography.
Further, it is assumed that its width can be approximated by a relation $w \propto Q^\alpha$ (Equation~\ref{eq:wcf}) with $\alpha\in [0,1]$; typically $\alpha \approx 1/2$ (a so-called regime channel).
End members $\alpha=0$ or $\alpha=1$ correspond a  subaerial channels of constant width (a slot canyon) or depth (no natural equivalent), respectively.
% 
For a steady state R-channel, it is assumed that  $\Psi$ is constant (approximated by the gradient of the Shreve \citeyear{shreve1972} potential and that $S$ adjusts in steady state with $\Psi$ and $Q$.
Note that the R-channel model presented above in Section~\ref{sect:sub_mode} Equation~\ref{eq:dS_dt} calculates $\Psi$ from the time-evolving $S_g$ via the Darcy-Weisbach Equation~\ref{eq:dh} and thus no Shreve approximation is then needed.
% Conversely, pipe flow has $S$ fixed and $\Psi$ adjusts to satisfy the given $Q$.
% 
Pipe flow-like conditions occur when an R-channel is subjected to rapid discharge variations such that the channel cannot adjust its size.
In this case, it is assumed that the cross-sectional area $S$ is fixed, and $\Psi$ adjusts to the specified $Q$.
For both R-channel and pipe flow, it is assumed that $D_h \propto S^{1/2}$.

With these assumptions, the Darcy-Weisbach equation~\eqref{eq:DW} can be solved for the not-fixed quantity: $H$ for a subaerial channel, $S$ for a R-channel, and $\Psi$ for pipe-flow.
Then, using equations~\eqref{eq:tau-omega}, the shear stress and stream power can be calculated.
These results of are summarised in Table~\ref{tab:eqs1}.
Now width integrated shear stress and sediment transport can be calculated for all regimes and all transport formulas.
The results of the 12 combinations are presented in Table~\ref{tab:Qs} and also in Table~S1 where the fractions in the exponents are approximately given by decimal numbers for ease of comparison.

@@

Table~\ref{tab:Qs} shows how the proxy $w\tau$ as well as $Q_s$ of the MPM, EH, and Bagnold sediment transport formulas scale with respect to $Q$, $\Psi$ or $S$ for the three different regimes and for different $\alpha$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow regime.
The total transport formula EH gives a slightly stronger dependence on all variables as should be expected due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$  for the others.
Albeit, the sediment transport response in pipe flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxies $w_g\,\tau_g$ \& $w_f\,\tau_f$ used below scale only as $Q^2$ for pipe-flow, whereas the transport scales at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts the relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{2}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly greater than the other transport relations.

The sediment discharge capacity in the steady state R-channel scales very similarly to the subaerial channel case for all relations and virtually identically for the $\alpha=0.5$ case.
However, the head gradients $\Psi$ are likely higher for comparable $Q$ in an ice-sheet marginal or alpine glacier setting than in a subaerial channel \citep{alley1997}.
Thus, sediment transport capacity is likely much higher in a steady-state R-channel.

@@

>>>>

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom.
  This quantity varies with water velocity and channel width.
  In subaerial channels,  sediment transport capacity can evolve consistently because water discharge variations are accommodated by changes in flow depth, width, and water velocity.
  By contrast, subglacial channels grow at a slower rate than most water discharge variations, causing a variable relationship between channel shape, water velocity, and discharge.
  Using numerical models applied to two observed hydrographs, we evaluate the differences in sediment transport capacity variability between subaerial and subglacial channels.
  The numerical experiments confirm that subglacial sediment transport is highly non-linear with respect to water discharge.
  Yet, subglacial sediment transport capacity can vary less if water discharge variations are closer to equilibrium with subglacial channel size.
  In a dedicated numerical experiment, we show that the changing channel size results in sediment transport capacity peaking before the maximum water discharge.
  Results also indicate that high subglacial sediment transport capacities can occur across a wide range of water discharges.
  This implies that hysteresis in channel size causes a variable relationship between sediment and water discharge in a transport-limited subglacial system.
  The implications are discussed for sediment discharge from glaciers with different hydro-climatic forcings.
  We also evaluate the conditions under which subglacial sediment transport capacity could co-vary with water discharge.
  These findings can also improve interpretations of sediment discharge records in glacierized catchments.
\end{abstract}

@@

This manuscript has two objectives: 1) to evaluate the variability of subglacial sediment transport capacity with evolving subglacial channel size and different hydrographs and 2) to establish whether sub-seasonal water discharge can co-vary with sediment transport capacity in subglacial systems.
We use numerical models to examine sediment transport capacity in both subglacial channels that evolve in size and subaerial channels that remain fixed.
They are applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and a land-terminating glacier in Greenland (Leverett Glacier).
We run a model ensemble to evaluate the sediment transport capacity variability in subglacial channels compared to subaerial channels.
Model outputs demonstrate the specific processes that can drive variability in sediment discharge capacity in subglacial systems.
These findings indicate differences in the relationships amongst water discharge, channel geometry, and water velocity in subglacial and subaerial channels.
The manuscript then discusses the implications of variability in sediment transport capacity from glaciers and the interpretation of sediment transport records.

@@

\begin{table}[hbt!]
  \centering
  \caption{Variables, parameters, and constants used in this work.
    Where two values are given, the first refers to  \alpine{}, a scenario from Fieschergletscher, and the second to a glacier marginal to the \icesheet{} a scenario from Leverett Glacier.
    A second line refers to the range of values examined in the parameter search.}
  \small 
  \begin{tabular}{ l  c  c c }
    Name &Symbol&  Value&Units \\
         && (\alpine{} or \icesheet{})\\
    \hline
    \textbf{Variables}  & & & \\
    Water discharge  & $Q$& & $\mathrm{m^{3}\,s^{-1}}$ \\
    Water velocity (subglacial, subaerial)  & $v$, ($v_g,\,v_{f}$)& & $\mathrm{m\,s^{-1}}$ \\
    Channel wetted area (subglacial, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
    Channel depth (subaerial) & $H$&& $\mathrm{m}$\\
    Hydraulic diameter &$D_h$&&$\mathrm{m}$\\
    Width of channel floor (subglacial, subaerial) & $w$, ($w_g,w_f$)&  & $\mathrm{m}$     \\
    Hydraulic head &$\Delta h$&& $\mathrm{m}$\\
    Hydraulic gradient &$\Psi=\frac{\Delta h}{l}$&& $\mathrm{m\, m^{-1}}$\\
    Shear stress (subglacial, subaerial) & $\tau$, ($\tau_g,\,\tau_f$) && $\mathrm{Pa \, m^{-2}}$ \\
x    Stream power & $\Omega$ && $\mathrm{ kg \, m\, s^{-3}}$ \\

@@

\section{Results}

@@

While transport-limited states at glaciers are likely rare \citep[e.g.][]{alley1997}, abundant sediment could persist underneath some glaciers \citep[e.g.][]{walter2014,stevens2022,delaney2022}.
As a result, sediment export can be hydrologically controlled and efficient \citep[e.g.][]{swift2005,andresen2024}.
At these glaciers, the great variability in subglacial sediment transport capacity may make it difficult to link sediment discharge to hydrology, especially when peak events occur \citep{cowan1988,delaney2018,lu2022}.
The results here suggest that different channel sizes with the same hydrology forcing can result in very different sediment transport capacities, making it challenging to establish the effects of individual extreme precipitation or melt events.

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.7\linewidth]{Fig7.pdf}
  \caption{Response of sediment transport capacity (orange line) and channel size (bottom row) to a water discharge event under different channel conditions.
    a) Pipe flow with fixed channel size described in Section~\ref{sect:scaling}.
    b) R-channel described in Section \ref{sect:sub_mode}.
    c) Subaerial channel described in Section~\ref{sect:fluv}.
    d) Steady state R-channel (channel size evolves in equilibrium with water discharge) described in Section~\ref{sect:scaling}.
    Note axes not to scale.
    Gyres on bottom panels represent shear stress on channel walls.
    $t_1$ ($t_2$) represents low (high) water discharge. 
  }
  \label{fig:chan_types}
\end{figure}

Catchments with reduced water discharge variability may experience yet less variability in sediment transport capacity, shown by the \alpine{} and \icesheet{} model outputs.
Indeed, the more decoupled and sporadic relationship between model outputs and water discharge in \alpine{} results from the relatively larger discharge variations on sub-daily to weekly timescales (Figure~\ref{fig:Qw_vari}).
This occurs as the subglacial channel's size evolves slowly compared to the variations in water discharge (Figure~\ref{fig:chan_types} b).
High water discharge variability in \alpine{} may cause shear stress to approach $Q^{2}$ in assuming pipe-flow conditions where the water discharge varies largely compared to channel size \citep[Figure~\ref{fig:chan_types}; Section \ref{sect:scaling}; c.f.][]{alley1997}.
Conversely, the reduced relative variability in water discharge in \icesheet{} and stronger correlation between width-integrated shear stress may result in the subglacial channel's size being closer to equilibrium from the smaller variations in water discharge (Figure~\ref{fig:chan_types} d).
In this case, the exponent on water discharge for shear stress is likely substantially less than $w\tau \propto Q^2$, but greater than $w \tau \propto Q^{\frac{4}{5}}$ that occurs in a steady-state R-channel (see Section \ref{sect:scaling}).
As a result, there is a stronger relationship between subglacial model outputs and water discharge in \icesheet{} (Figure~\ref{fig:Qw_vari}) and less variability model outputs in the \icesheet case (Figure~\ref{fig:multi_run}).

@@

The co-varying relationship between sediment discharge capacity and water discharge in subaerial conditions could be noticeable $\sim20$\,\unit{km} downstream of the Leverett site.
A strong correlation persists between sediment plume size and the Watson River's water discharge into the Kangerlussuaq fjord \citep[Figure~\ref{fig:chan_types}\,c][]{chu2009,mcgrath2010}.
In contrast, in marine-terminating glacier catchments, a less consistent relationship may occur between water discharge or melt extent and sediment plume size \citep{chu2012,tedstone2012}.
Here, ocean water pressurizes subglacial channels at the ice front \citep[e.g.][]{how2017}, so the observed reduced correlation could result from the inconsistent relationship between subglacial sediment transport capacity and water discharge (Figures~\ref{fig:Qw_vari} and \ref{fig:chan_types}\,b).

Water discharge measurements at less than daily timescales could severely limit subglacial sediment transport models in capturing specific events.
The decreased model output variability beyond $1$--$5$ days of smoothing could result in water discharge being in equilibrium with the subglacial channel when they are in fact not (Figure~\ref{fig:multi_run}).
Over this period, water discharge variations could lead to a stronger relationship with sediment discharge capacity (Figure~\ref{fig:chan_types} d and Section ~\ref{sect:scaling}).
Particularly, if water discharge evolves slowly with respect to the subglacial channel size, then it will covary with sediment transport capacity (Figure~\ref{fig:chan_types} d).
Such a strong relationship would not represent the impact of actual shorter-term fluctuations in water discharge.
Note, in this case, that if  pipe flow is assumed, with pressurized flow, but no channel evolution, then  sediment discharge capacity would covary with water discharge (Figure~\ref{fig:chan_types} a).

Clockwise hysteresis loops between sediment concentration and water discharge suggest that sediment availability is reduced over the event scale \citep[][]{williams1989}.
These loops are often observed in glacierized catchments and used to suggest that access to subglacial sediment has been limited \citep[e.g.][]{collins1979,willis1996,richards2003,stott2007,delaney2018}.
Results here suggest that clockwise hysteresis could also be expected in transport-limited regimes in many subglacial environments.
Here, the smaller channel size on the rising limb would result in greater sediment transport capacity and sediment uptakes (Figure~\ref{fig:chan_types} b).
A larger channel size would reduce sediment transport capacity for an equivalent water discharge on the falling limb, creating clockwise hysteresis.

@@

The manuscript's first objective aims to evaluate the relative variability in sediment transport capacity between the two channel types.
Results demonstrate that sediment transport capacity variability is higher in subglacial channels with evolving size across a range of channel conditions.
Increased variability occurs even in environments where the water discharge has relatively small diurnal variations in the \icesheet case.
Yet, elevated variability is especially strong with highly variable water discharge in the \alpine case.
Further evaluation is needed to establish the role of highly variability in sediment transport capacity in evaluating hydro-climatic signals from sediment records.
Greater variability in sediment transport capacity may also lead to sediment exhaustion by subglacial water repeatedly crossing the mobilization threshold across a range of water discharges.

The manuscript's second objective is to establish if water discharge varies with subglacial sediment transport capacity.
Results here suggest that, even in a transport-limited subglacial system, an incoherent relationship between water and subglacial sediment discharge could be expected in most systems.
This presents a challenge in linking hydro-climatic conditions or events to sediment export from glaciers.
In subglacial channels, the timing of peak sediment discharge capacity occurs before peak water discharge during an event, due to evolving channel size.
In subaerial channels, the timing of peak sediment discharge capacity and water discharge coincide.
Covariance between water discharge and sediment transport capacity could be possible when water discharge varies at a slower rate than subglacial channel size.
In natural environments, this appears to rarely be the case, except where basal melt dominates.

@@

\authorcontribution{
  I.D. designed the study, developed, and implemented the model and experiments, and wrote the manuscript.
  A.J.T. helped interpret findings from Leverett Glacier and guided experiment design.
  M.A.W. and D.F. provided data from Fieschergletscher and contributed to analysis.
  All authors provided key inputs in writing and editing the manuscript.} %% this section is mandatory

@@

Here three kinds of drainage regimes are examined, namely, pipe-flow (i.e. R-channels which do not have time to adjust their size to discharge conditions; Figure~\ref{fig:chan_types} a), subaerial channels (Figure~\ref{fig:chan_types} c), steady-state R-channels \citep[Figure~\ref{fig:chan_types} d]{rothlisberger1972}.
Here we show the sediment transport capacity scaling with a given water discharge and hydraulic gradient for three different sediment transport capacity formulas: Meyer-Peter M\"uller  \citep[MPM; ][]{meyer1948}, Engelund and Hansen \citep[EH; ][]{engelund1967}, and Bagnold \citep{bagnold1980}; additionally width-integrated shear stress is assessed, as used as proxy for sediment transport throughout the manuscript.

Sediment discharge is given by the MPM, EH, and Bagnold formulations as
\begin{equation}
  \label{eq:Qs_eq}
  Q_s \propto w\, \tau^{3/2}, \quad Q_s \propto w\, \tau^{5/2}, \quad Q_s \propto w \left(\frac{\Omega}{w}\right)^{3/2} H^{-2/3},
\end{equation}
respectively, for conditions well above the transport onset threshold.

@@

As in Section~\ref{sect:fluv}, the subaerial channel is assumed to have a width $w$ much greater than its depth $H$, such that $D_h\approx 4h$, and to have a constant head gradient ($\Psi$) given by the topography.
Further, it is assumed that its width can be approximated by a relation $w \propto Q^\alpha$ (Equation~\ref{eq:wcf}) with $\alpha\in [0,1]$.
End members $\alpha=0$ or $\alpha=1$ correspond a  subaerial channels of constant width (a slot canyon) or depth (no natural equivalent), respectively.
% 
For a steady-state R-channel, it is assumed that  $\Psi$ is constant (approximated by the gradient of the Shreve \citeyear{shreve1972} potential and that $S$ adjusts in steady state with $\Psi$ and $Q$.
Note that the R-channel model presented above in Section~\ref{sect:sub_mode}, Equation~\ref{eq:dS_dt} calculates $\Psi$ from the time-evolving $S_g$ via the Darcy-Weisbach Equation~\ref{eq:dh} and thus no Shreve approximation is then needed.
% Conversely, pipe flow has $S$ fixed and $\Psi$ adjusts to satisfy the given $Q$.
% 
Pipe flow-like conditions occur when an R-channel is subjected to rapid discharge variations such that the channel cannot adjust its size.
In this case, it is assumed that the cross-sectional area $S$ is fixed, and $\Psi$ adjusts to the specified $Q$.
For both steady-state R-channel and pipe flow, it is assumed that $D_h \propto S^{1/2}$.

With these assumptions, the Darcy-Weisbach equation~\eqref{eq:DW} can be solved for the not-fixed quantity: $H$ for a subaerial channel, $S$ for a R-channel, and $\Psi$ for pipe-flow.
Then, using equations~\eqref{eq:tau-omega}, the shear stress and stream power can be calculated.
These results of are summarised in Table~\ref{tab:eqs1}.
Now width integrated shear stress and sediment transport can be calculated for all regimes and all transport formulas.
The results of the 12 combinations are presented in Table~\ref{tab:Qs} and also in Table~\ref{tab:Qs2} where the fractions in the exponents are approximately given by decimal numbers for ease of comparison.

@@

Table~\ref{tab:Qs} shows the scaling of the proxy $w\tau$ as well as $Q_s$ of the MPM, EH, and Bagnold sediment transport formulas with respect to $Q$, $\Psi$ or $S$ for the three different regimes and for different $\alpha$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow regime.
The total transport formula EH gives a slightly stronger dependence on all variables due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$  for the others (Equation~\ref{eq:Qs_eq}).
Albeit, the sediment transport response in pipe flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxies $w_g\,\tau_g$ and $w_f\,\tau_f$ used below scale only as $Q^2$ for pipe-flow, whereas the transport scales at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts the relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{3}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly greater than the other transport relations.

The sediment discharge capacity in the steady state R-channel scales very similarly to the subaerial channel case for all relations and virtually identically for the $\alpha=0.5$ case (Table\ref{tab:eqs1}).
However, the head gradients $\Psi$ are likely higher for comparable $Q$ in an ice-sheet marginal or alpine glacier setting than in a subaerial channel \citep{alley1997}.
Thus, sediment transport capacity is likely higher in a steady-state R-channel.

@@

\begin{table}
  \caption{As Table~3 but with the fractional exponents stated as rounded decimal numbers.  For the subaerial regime the exponents are for displayed for the likely width-exponent $\alpha=1$, as well as its end members 0 (slot canyon) and 1 (only width increases and not depth).
  }
  \small
  \label{tab:Qs2}
  \begin{tabular}{lllll}
    & Width \(\times \,\, \tau\) & MPM & EH & Bagnold\\
    & \(w\, \tau\) & \(Q_s \propto w\, \tau^{3/2}\) & \(Q_s \propto w\, \tau^{5/2}\) & \(Q_s \propto w^{-1/2}\, \Omega^{3/2} H^{-2/3}\)\\
    \hline
    Subaerial (\(\alpha=0\)) & \(f^{0.7}\, Q^{0.7}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{1.7} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{1.1} \, \Psi^{1.7}\)\\
    Subaerial (\(\alpha=0.3\)) & \(f^{0.7}\, Q^{0.8}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{1.3} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{1.0} \, \Psi^{1.7}\)\\
    %Subaerial (\(\alpha=0.5\)) & \(f^{0.7}\, Q^{0.8}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{1.3} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{1.0} \, \Psi^{1.7}\)\\
    Subaerial (\(\alpha=1\)) & \(f^{0.7}\, Q^{1.0}\,  \Psi^{0.7}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.8}\, Q^{0.7} \, \Psi^{1.7}\) & \(f^{-0.2}\, Q^{1.0} \, \Psi^{1.7}\)\\[3pt]
    R-channel & \(f^{0.4}\, Q^{0.8} \, \Psi^{0.6}\) & \(f^{0.5}\, Q \,\,\,\, \Psi\) & \(f^{0.7}\, Q^{1.4}\, \Psi^{1.8}\) & \(f^{-0.2}\, Q^{1.0} \, \Psi^{1.7}\)\\
    Pipe & \(f \,\quad Q^{2\phantom{.0}} \, S^{-1}\) & \(f^{1.5}\, Q^3 \, S^{-2.5}\) & \(f^{2.5}\, Q^{5\phantom{.0}}\, S^{-4.5}\) & \(f^{1.5} \,\,\,\,\, Q^{4.5} \, S^{-4.7}\)\\
  \end{tabular}
\end{table}
